Curvature approximation on surfaces -- A Discrete Exterior Calculus Approach

### 2015-01-12 article/main.tex
article update, start cdec

  \begin{figure}
    \centering\includegraphics[width=.45\textwidth]{bilder/bunnyCurvature.jpg}
    \centering\includegraphics[width=.45\textwidth]{bilder/horseCurvature.jpg}
    \centering\includegraphics[width=.45\textwidth]{bilder/colourBar.jpg}
    \caption{Surface curvature approximations of the Stanford Bunny and a horse.
             Left: Gaussian curvature \( K \) (square root scaled colours).
             Right: Mean curvature \( H \) (linear scaled colours)}
    \label{figBunnyHorseCurvature}
  \end{figure}

@@

\section{Curvature approximations}
  \subsection{Curvature vector} \label{secCurvatureVector}
    The curvature vector \( \vec{H} \) is given by
    \begin{align}\label{eqMeanCurvCurvVec}
      \vec{H} = 2 H \vec{\nu}
    \end{align}
    for the mean curvature \( H \) and normals \( \vec{\nu} \). 
    In this paper, the mean curvature is set as the arithmetic mean of the two principal curvatures and
    \( H \) should be all over positive for a sphere (i.e. \( H\equiv1 \) on \( \mathds{S}^{2} \)).
    \( H \) and \( \vec{H} \) are extrinsic curvature values.
    This means, that an two dimensional ``inhabitant'' of the surface can't determine this kind of curvature 
    in contrast to the Gaussian curvatures.
    With the inclusion map \( \vec{x}:= \R^{3}\supset M \hookrightarrow \R^{3} \), which is the position vector of \( M \) in \( \R^{3} \),
    the curvature vector can be calculated by (see \cite[Ch. 5]{chen}, \cite[Ch. 4.5]{flanders} or \cite{elliott})
    \begin{align} \label{eqCurvatureVector}
      -\Delta_{B}\vec{x} = \vec{H} \formpunkt
    \end{align}
    (Note that \( \vec{x} \) is an isometric immersion with the common induced metric from \( \R^{3} \).)

@@

    On the abstract Voronoi cell \( \pi(\star v) \) (arise by projecting to the manifold) of a vertex \( v \), 
    it holds the Gauss-Bonnet theorem \cite{berger}:
    \begin{align}
      \label{eqGaussBonnet}
      \int_{\pi(\star v)}K \mu = 2\pi - \sum_{i=1}^{m_{v}}\tilde{\beta_{i}} - \int_{\partial\pi(\star v)} k_{g} ds \formpunkt
    \end{align}
    \( \mu\in\Omega^{2}(M)  \) is the volume form, \( k_{g} \) the geodetic curvature on the boundary edges of the cell 
    and \( \tilde{\beta_{i}} \) the \( m_{v} \) exterior angles of the boundary on the Voronoi vertices.
    This can be approximated on the simplicial complex \( K \) by omitting the projection.
    The geodetic curvature \( k_{g} \) vanish on the straight lines and on the bucklings, where the dual edges intersect the primal edges.
    Hence, in our DEC notation, we get
    \begin{align}
      \label{eqDisGaussBonnet}
      \left\langle *K_{GB} , \star v \right\rangle &:= 2\pi - \sum_{i=1}^{m_{v}}\beta_{i}
    \end{align}
    on a vertex \( v\in K \), with the exterior angles \( \beta_{i} \).
    \begin{figure}
      \begin{minipage}[htp]{.235\textwidth}
        \centering\includegraphics[width=.99\textwidth]{bilder/tikz/GaussBonnet.eps}
      \end{minipage}
      \hfill
      \begin{minipage}[htp]{.235\textwidth}
        \centering\includegraphics[width=.99\textwidth]{bilder/tikz/ElementGaussBonnet.eps}
      \end{minipage}
      \caption{Left: \(  m = m_{v} =: \tilde{m}+1 \) exterior angles \( \beta_{i} \) on the vertices of a Voronoi cell.
                The solid lines (with orientations) marks the dual (Voronoi) edges.
               Right: Angles in a primal triangle referring to the edges from the circumcenter subdivision.}
      \label{figGaussBonnet}
    \end{figure}
    We see in figure \ref{figGaussBonnet}, that it is possible to calculate the exterior angles by the local interior angles 
    \( \alpha_{0} \) and \( \alpha_{1} \) on \( v \) for a
    triangle \( \sigma^{2}\succ v \):
    \begin{align}
      \beta_{\bullet} = \pi - \gamma_{0} - \gamma_{1} = \alpha_{0} + \alpha_{1} \formpunkt
    \end{align}
    Because the interior angles are determinable by the length measures of the edges, we get 
    \begin{align}
      \sum_{i=1}^{m_{v}}\beta_{i} &= \sum_{\sigma^{2}\succ v} \sum_{\sigma^{2}\succ\sigma^{1}\succ v}
                               \text{atan2}\left( 2\left|\star\sigma^{1} \cap \sigma^{2} \right|, 
                                 \left|\sigma^{1}\right|\right)
    \end{align}
    for the angle sum in \eqref{eqDisGaussBonnet}.

@@

  \begin{align}
    \label{eqSphere}
    \varphi\left( x,y,z \right) &:= x^{2} + y^{2} + z^{2} - 1 \\
    \label{eqEllipsoid}
    \varphi\left( x,y,z \right) &:= \left( 3x \right)^{2} + \left( 6y \right)^{2} + \left( 2z \right)^{2} - 9\\
    \label{eqQuartic}
    \varphi\left( x,y,z \right) &:= \left( x - z^{2} \right)^{2} + \left( y - z^{2} \right)^{2} + z^{2} - 1
  \end{align}

@@

  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/ellipsoid/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/ellipsoid/H.jpg}
    \end{minipage}
    \caption{Ellipsoid \eqref{eqEllipsoid}: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figEllipsoid}
  \end{figure}

@@

  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/quartic/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/quartic/H.jpg}
    \end{minipage}
    \caption{Quartic surface\eqref{eqQuartic}: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figQuartic}
  \end{figure}

@@

  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/torus/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/torus/H.jpg}
    \end{minipage}
    \caption{Torus: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figTorus}
  \end{figure}

  \begin{figure}
    \centering\includegraphics[width=.49\textwidth]{bilder/torus/L2K.eps}
    \centering\includegraphics[width=.49\textwidth]{bilder/torus/L2H.eps}
    \caption{Torus: Relative \( L_{2} \) error for \( K \) (top) and
                                                     \( H \) (bottom) for different \( h \) in a log-log-plot.}
    \label{figTorusL2Err}
  \end{figure}

>>>>

  \begin{figure}
    \centering\includegraphics[width=.45\textwidth]{bilder/bunnyCurvature.jpg}
    \centering\includegraphics[width=.45\textwidth]{bilder/horseCurvature.jpg}
    \centering\includegraphics[width=.45\textwidth]{bilder/colourBar.jpg}
    \caption{Surface curvature approximations of the Stanford Bunny (39998 DOFs) and a horse (89983 DOFs).
             Left: Gaussian curvature \( K \) (square root scaled colours).
             Right: Mean curvature \( H \) (linear scaled colours)}
    \label{figBunnyHorseCurvature}
  \end{figure}

@@

\section{Curvature approximations} \label{secCurvatureApprox}
  \subsection{Curvature vector} \label{secCurvatureVector}
    The curvature vector \( \vec{H} \) is given by
    \begin{align}\label{eqMeanCurvCurvVec}
      \vec{H} = 2 H \vec{\nu}
    \end{align}
    for the mean curvature \( H \) and normals \( \vec{\nu} \). 
    In this paper, the mean curvature is set as the arithmetic mean of the two principal curvatures and
    \( H \) should be all over positive for a sphere (i.e. \( H\equiv1 \) on \( \mathds{S}^{2} \)).
    \( H \) and \( \vec{H} \) are extrinsic curvature values.
    This means, that an two dimensional ``inhabitant'' of the surface can't determine this kind of curvature 
    in contrast to the Gaussian curvatures.
    With the inclusion map \( \vec{x}:= \R^{3}\supset M \hookrightarrow \R^{3} \), which is the position vector of \( M \) in \( \R^{3} \),
    the curvature vector can be calculated by (see \cite[Ch. 5]{chen}, \cite[Ch. 4.5]{flanders} or \cite{elliott})
    \begin{align} \label{eqCurvatureVector}
      -\Delta_{B}\vec{x} = \vec{H} \formpunkt
    \end{align}
    (Note that \( \vec{x} \) is an isometric immersion with the common induced metric from \( \R^{3} \).)

@@

    On the abstract Voronoi cell \( \pi(\star v) \) (arise by projecting to the manifold) of a vertex \( v \), 
    it holds the Gauss-Bonnet theorem \cite{berger}:
    \begin{align}
      \label{eqGaussBonnet}
      \int_{\pi(\star v)}K \mu = 2\pi - \sum_{i=1}^{m_{v}}\tilde{\beta_{i}} - \int_{\partial\pi(\star v)} k_{g} ds \formpunkt
    \end{align}
    \( \mu\in\Omega^{2}(M)  \) is the volume form, \( k_{g} \) the geodetic curvature on the boundary edges of the cell 
    and \( \tilde{\beta_{i}} \) the \( m_{v} \) exterior angles of the boundary on the Voronoi vertices.
    This can be approximated on the simplicial complex \( K \) by omitting the projection.
    The geodetic curvature \( k_{g} \) vanish on the straight lines and on the bucklings, where the dual edges intersect the primal edges
    in right angles.
    Hence, in our DEC notation, we get
    \begin{align}
      \label{eqDisGaussBonnet}
      \left\langle *K_{GB} , \star v \right\rangle &:= 2\pi - \sum_{i=1}^{m_{v}}\beta_{i}
    \end{align}
    on a vertex \( v\in K \), with the exterior angles \( \beta_{i} \).
    \begin{figure}
      \begin{minipage}[htp]{.235\textwidth}
        \centering\includegraphics[width=.99\textwidth]{bilder/tikz/GaussBonnet.eps}
      \end{minipage}
      \hfill
      \begin{minipage}[htp]{.235\textwidth}
        \centering\includegraphics[width=.99\textwidth]{bilder/tikz/ElementGaussBonnet.eps}
      \end{minipage}
      \caption{Left: \(  m = m_{v} =: \tilde{m}+1 \) exterior angles \( \beta_{i} \) on the vertices of a Voronoi cell.
                The solid lines (with orientations) marks the dual (Voronoi) edges.
               Right: Local angles (\( \beta^{\sigma^{2}}:=\beta_{i} \)) in a primal triangle \( \sigma^{2} \) referring to the edges of the circumcenter subdivision.}
      \label{figGaussBonnet}
    \end{figure}
    We see in figure \ref{figGaussBonnet}, that it is possible to calculate the exterior angles by the local interior angles 
    \( \alpha_{0} \) and \( \alpha_{1} \) on \( v \) for a
    triangle \( \sigma^{2}\succ v \):
    \begin{align}
      \beta^{\sigma^{2}} = \pi - \gamma_{0} - \gamma_{1} = \alpha_{0} + \alpha_{1} \formpunkt
    \end{align}
    Because the interior angles are determinable by the length measures of the edges, we get 
    \begin{align}
      \sum_{i=1}^{m_{v}}\beta_{i} &= \sum_{\sigma^{2}\succ v} \sum_{\sigma^{2}\succ\sigma^{1}\succ v}
                               \text{atan2}\left( 2\left|\star\sigma^{1} \cap \sigma^{2} \right|, 
                                 \left|\sigma^{1}\right|\right)
    \end{align}
    for the angle sum in \eqref{eqDisGaussBonnet}.

@@

  \subsection{Test surfaces}
    For our numerical tests, we use the following 4 surfaces, with analytical known Gauss and mean curvatures:

    \begin{description}
      \item[Sphere] (unit) with given signed distance function  
        \begin{align}
             \label{eqSphere}
             \varphi\left( x,y,z \right) &:= x^{2} + y^{2} + z^{2} - 1 \formpunkt
        \end{align}
        The Gauss and mean curvature are all over 1 on the unit sphere.
        \item[Ellipsoid] (tri-axial; length of semi-principal axes: \( \left( 1,0.5,1.5 \right) \)) 
            with given signed distance function  
              \begin{align}
                \label{eqEllipsoid}
                \varphi\left( x,y,z \right) &:= \left( 3x \right)^{2} + \left( 6y \right)^{2} + \left( 2z \right)^{2} - 9
              \end{align}
            and curvatures (see Figure \ref{figEllipsoid})
            \begin{align}
              K(x,y,z) &= \frac{11664}{(81+972y^{2}-20z^{2})^{2}} \\
              H(x,y,z) &= \frac{36 \left(45+54 y^2-10 z^2\right)}{\left(81+972 y^2-20 z^2\right)^{3/2}} \formpunkt
            \end{align}
  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/ellipsoid/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/ellipsoid/H.jpg}
    \end{minipage}
    \caption{Ellipsoid \eqref{eqEllipsoid}: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figEllipsoid}
  \end{figure}

       \item[Quartic surface] with given signed distance function
            \begin{align}
              \label{eqQuartic}
              \varphi\left( x,y,z \right) &:= \left( x - z^{2} \right)^{2} + \left( y - z^{2} \right)^{2} + z^{2} - 1
            \end{align}
            and curvatures (see Figure \ref{figQuartic})
            {\footnotesize
            \begin{align}
              K(x,y,z) &= -\frac{1 + 2x - 2x^{2} + 2y - 2y^{2} + 2(-3 + x + y)z^{2}}
                          {1 - 4(-2 + x + x^{2} + y - 2xy + y^{2})z^{2}} \\
              H(x,y,z) &= -\frac{x - x^{2} + y - y^{2} + (-7 + 3x + 3y)z^{2}}
                         {1 - 4(-2 + x + x^{2} + y - 2xy + y^{2})z^{2}}\formpunkt
            \end{align}
            }
  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/quartic/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/quartic/H.jpg}
    \end{minipage}
    \caption{Quartic surface\eqref{eqQuartic}: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figQuartic}
  \end{figure}

      \item[Torus] with major radius\footnote{distance from the center of the tube to the center of the torus} \( R = 2 \)
                   and minor radius\footnote{radius of the tube} \mbox{\( r=0.5 \)}
                   with given signed distance function  
              \begin{align}
                \label{eqTorus}
                \varphi\left( x,y,z \right) &:= \left( 2 - \sqrt{x^{2}-y^{2}} \right)^{2} + z^{2} - 0.25 
              \end{align}
                  and curvatures (see Figure \ref{figTorus})
              \begin{align}
                K(x,y,z) &= 4 - \frac{8}{\sqrt{x^{2}+y^{2}}} \\
                H(x,y,z) &= 2 - \frac{2}{\sqrt{x^{2}+y^{2}}} \formpunkt
              \end{align}
  \begin{figure}
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/torus/K.jpg}
    \end{minipage}\hfill
    \begin{minipage}[htp]{.23\textwidth}
      \centering
      \includegraphics[width=0.99\textwidth]{bilder/torus/H.jpg}
    \end{minipage}
    \caption{Torus \eqref{eqTorus}: Gaussian curvature \( K \) (left) and mean curvature \( H \) (right).}
    \label{figTorus}
  \end{figure}
    \end{description}

  \subsubsection{Notes on meshing}
    Except for the torus, all macro triangulations are icosahedrons.
    In \cite{arakawa}, we see that the 20 plane faces are able to divide in \( N^{2} \) equal equilateral triangles
    and all vertices can projected to the smooth surface.
    Unfortunately, this procedure doesn't results in a well-centered simplicial complex for the ellipsoid and the quartic surface.
    Our resort for this problem provide a mechanical evolution for the coordinates \( \vec{x}_{i} \) of the vertices \( v_{i}\in K \)
    \begin{align}\label{eqForceLaw}
      \frac{\mathrm{d} \vec{x}_{i}}{\mathrm{d} t} &= \vec{F}(\vec{x}_{i}) \formpunkt
    \end{align}
    The force vector \( \vec{F}(\vec{x}_{i}) \) describe the interaction with the adjoining edges \( \sigma^{1}\succ v_{i} \) and
    triangles \( \sigma^{2}\succ v_{i} \).
    We use two approaches for \( \vec{F} \), who will linear combinated.
    One is to displace the vertices, so that all edges grasp at the same length.
    This is similar to a overlapping sphere (OS)/ Voronoi tesselation (VT) hybrid model in \cite{meshCooper}, with fixed connectivity, 
    for simulating the arrangement of cells in tissue.
    The other approach is to choose \( \vec{F} \) so, that the edges \( \sigma^{1}\succ v_{i} \) will become longer or shorter in dependency
    of the enclosed angle of \( \sigma^{1} \) and the other edge \( \tilde{\sigma}^{1}\prec\sigma^{2} \) 
    on all triangles \( \sigma^{2}\succ v_{i} \).
    The aspired optimum (which can't be reached in general) can be a enclosed angle of \( \frac{\pi}{3} \) for example.

    The equation \eqref{eqForceLaw} can be numerical solved by the explicit Euler method.
    It must be pointed out, that the smooth surface \( M \) gives us a geometrical constrain for the coordinates \( \vec{x}_{i} \).
    That's why we project all vertices to \( M \) after every Euler step.
    A explicit base method of higher order doesn't result in faster convergence in practice 
    because of the projection
    and a implicit method is hard to implement with all of the dependencies of \( \vec{F} \) in mind.

    The torus mesh in figure \ref{figTorus} can be obtain by a uniform triangulation of a tube and clue the both ends together 
    with a twist of \( 4\pi \) along the tube.
    The meshes for the bunny and the horse in figure \ref{figBunnyHorseCurvature} are produced with the  
    great AVCD\footnote{Approximated Centroidal Voronoi Diagram} remeshing tool (see \cite{valette}) with some handmade corrections 
    (clearing up the few \mh{1}{rings} with less than 5 triangles).

   \subsection{Comparing method}

    For the three test surfaces of genus 0 (sphere, ellipsoid, quartic surface) exists experimental results in \cite{heine} by C-J. Heine. 
    There is a isoparametric finite element approach deployed of degree 1 till 4 for the approximation of the curvatures.
    We only want to use the results of the degree 2 FE-Method (FEM-D2) to compare with our calculations,
    because the linear FEM-D1, which is the only really comparable with regard to the computation effort, is not convergent
    and FEM-D3, resp. FEM-D4, is too expensive in contrast to our methods.
    The lack of convergence of the discrete curvature vector with a linear FEM is discussed in \cite{nochetto}.
    The advantage of a FE approach is that the triangulation doesn't need to be well-centered.
    Therefor, the global mesh refinement strategy in \cite{heine} is to bisect the longest edge of a triangle.
    This produce triangulations of \mh{1}{rings} with 4 or 8 elements in about equal parts,
    i.e. the average are approximately  \mh{1}{rings} with 6 elements like our triangulations
    (all \mh{1}{rings} have 6 elements, except the 12 defects with 5 triangles).
    That's the reason, why the number of local DOFs (lDOFs) per triangle is a plausible measurement for the relative costs
    to compare the FE and DEC methods.
    E.g. for one value on a global vertex with 6 adjacent elements and a linear method 
    (i.e. local coefficients matrix in \( \R^{3 \times 3} \)) we get a row with 7 NNZ\footnote{Number of NonZero elements}
    in the assembled global coefficients matrix independently of the choice of the method. 
    For the calculation of the DEC-discretized curvature vector (LX) with \eqref{eqDisCurvatureVector}, 
    we have to solve the Laplace-Beltrami for the 3 components of the position vector on every vertices.
    Therefor, there are 9 lDOFs on every triangle element.
    For the Weingarten map approach (W) with \eqref{eqDisWeingarten}, we must calculate a \mh{3 \times 3}{matrix}
    on a vertex (\( \Rightarrow  \) 27 lDOFs).
    If the normals are unknown (W,\( \avn \)), then 3 lDOFs are added per vertex (\( \Rightarrow  \) 36 lDOFs). 
    The Gauss-Bonnet approximation gives us only 3 DOFs per vertex (\( \Rightarrow \) 9 lDOFs).
    In the FE-approaches (FEM-D*) in \cite{heine} is the Weingarten map and the curvature vector, as interim result, to be calculated.
    But at this, is to respect the growing number of DOFs per value with higher degrees of the elements.
    A summary is displayed in table \ref{tabLDOFs}.
    \begin{table*}[tbp]
      \definecolor{mygray}{gray}{.75}
      \centering
      \begin{tabular}{|c|>{\columncolor{mygray}}c|>{\columncolor{mygray}}c|>{\columncolor{mygray}}c|>{\columncolor{mygray}}c|c|>{\columncolor{mygray}}c|c|c|}
        \hline
        Method & GB & LX & W & W,\( \avn \) & FEM-D1 & FEM-D2 & FEM-D3 & FEM-D4 \\\hline
        lDOFs  & 3  & 9  & 27& 36           & 36     & 72     & 120    & 180 \\\hline
      \end{tabular}
      \caption{lDOFs for different methods in ascending order.
               Our 4 appoaches and the FEM-D2 from \cite{heine} for comparing are highlighted gray}
      \label{tabLDOFs}
    \end{table*}

  \subsection{Numerical experiments}
    We have implemented all curvature approximation approaches of section \ref{secCurvatureApprox} in a AMDiS framework.
    So we defined only the (triangle) element coefficients matrix as element operators and used linear functions for the interpolation of
    the results.
    The solving of the resulting linear system was done by the BiCGStab2\footnote{BiConjugate Gradient Stabilized method with \( l=2 \)}.
    Like in \cite{heine}, the errors were measured in the discrete relative \( L_{r} \)-norm (\( r\in\left\{ 2,\infty \right\} \)), 
    i.e. the absolute errors were globally normalized by the norm of the analytical solutions:
    \begin{align}
      \text{Err}_{r} := \frac{\left\| C - C_{\text{Meth}} \right\|_{L_{r}(|K|)}}{ \left\| C \right\|_{L_{r}(|K|)}} \formpunkt
    \end{align}
    \( C\in\left\{ K,H \right\} \) is the curvature value and ``\( \text{Meth} \)'' the used method.
    For \( r=2 \) we plot the errors for our test surfaces in figures \ref{figSphereL2Err}, \ref{figEllipsoidL2Err}, \ref{figQuarticL2Err} and
    \ref{figTorusL2Err} in a log-scale on both axis.
    The axis of abscissas describe the fineness of the triangulation and is measured in
    \begin{align}
      h := 2\max_{\sigma^{2}\in K} \left\| \vec{x}\left( v^{\sigma^{2}} \right) - \vec{x}\left( \star\sigma^{2} \right) \right\|_{2}
      \formkomma
    \end{align}
    with an arbitrary vertex \( v^{\sigma^{2}}\prec\sigma^{2} \),
    i.e. \( h \) is the maximum over all circumscribed diameters.
    We see in table \ref{tabEOC} the EOCs (Experimental Order of Convergence)
    for the both lowest \( h \) in the appropriate norm.
    \begin{table*}[tbp]
      \centering
      \begin{tabular}{|*{17}{c|}} \hline
         & \multicolumn{4}{c|}{Sphere} 
         & \multicolumn{4}{c|}{Ellipsoid} 
         & \multicolumn{4}{c|}{Quartic surface} &
        \multicolumn{4}{c|}{Torus} \\\cline{2-17}
          EOC & \multicolumn{2}{c|}{K} & \multicolumn{2}{c|}{H}
              & \multicolumn{2}{c|}{K} & \multicolumn{2}{c|}{H}
              & \multicolumn{2}{c|}{K} & \multicolumn{2}{c|}{H}
              & \multicolumn{2}{c|}{K} & \multicolumn{2}{c|}{H} \\\cline{2-17}
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \)
              & \( L_{2} \) & \( L_{\infty} \) \\\hline
        GB(K) / LX(H) & 2.0 & 2.0 & 2.4 & 1.7 & 
                  1.4 & NC  & 1.9 & NC &
                  1.8 & 1.8 & 1.8 & 1.8 &
                  1.9 & 1.0 & 1.9 & 1.4 \\\hline
           W    & 2.0 & 2.0 & 2.0 & 2.0 &
                  2.0 & 2.0 & 2.0 & 2.0 &
                  1.7 & 1.8 & 1.8 & 1.8 &
                  2.0 & 2.1 & 2.0 & 2.0 \\\hline
   W,\( \avn \) & 1.4 & NC  & 1.4 & NC  &
                  1.9 & NC  & 2.0 & NC  &
                  1.6 & 1.6 & 1.7 & 1.6 &
                  2.0 & 2.1 & 2.0 & 2.0 \\\hline
        FEM-D2  & 2.4 & 2.0 & 2.4 & 2.0 &
                  1.3 & 1.7 & 1.1 & 1.3 & 
                  2.0 & 2.2 & 1.4 & 1.4 &&&&\\\hline
      \end{tabular}
      \caption{Experimental Order of Convergence (EOC) of both finest triangulations for the relative discrete \( L_{2} \)-
      and \( L_{\infty} \)-norm. Values lower than 0.1 are denoted as NC (No Convergence).}
      \label{tabEOC}
    \end{table*}
    The reason for the choice of the finest meshes is the sometimes poor resolution of coarser triangulations.
    E.g. we see, by the mean of the slope, no or only lower convergence at mounted \( h \) for the quartic surface in figure
    \ref{figQuarticL2Err},
    because the mesh don't picture good enough the two ridges, so that the geometrical error exceeds the error of the method.
    In general, it is often to observe, that there is a kind of ``transient phase'' with respect to the order of convergence.
    And the higher the ``complexity'' of the surface the longer is this phase.

    However, on the sphere and the ellipsoid stagnate the EOCs for the discrete maximum norm on finer triangulations.
    In figure \ref{figSphereErrView} and \ref{figEllipsoidErrView} (top) we see that maximum errors sticks on the defects of the hexagonal
    mesh structure for the \( W,\avn \) methods respectively for \( K_{GB} \).
    Figure \ref{figEllipsoidErrView} (bottom) point to a similar error behavior for calculation of the curvature vector on the ellipsoid,
    but there are larger errors on the outer vertices of the 1-rings too.

  \subsection{Conclusion}
    With regard to the errors and the computational effort, the Gauss-Bonnet-Operator for the Gauss curvature (\( K_{GB} \)) and the
    curvature vector with the DEC-discretized Laplace-Beltrami-Operator for the mean curvature (\( H_{LX} \)) 
    seems to be the best choice of method for our used
    examples.
    But the advantage of the calculation of the Weingarten map is the higher information content, 
    because \( W \) gives us the whole shape-operator and the both principal curvatures as consequence
    and hence \( W \) must be only one time calculated to get \( K \) and \( H \) simultaneous.

    Table \ref{tabEOC} leads to guess of a nearly second order method for the DEC-discretized Weingarten map (\( W \)) if the normals of the smooth
    surface are known on every vertex.
    This partially not hold if the normals are averaged over the adjacent element normals (\( W,\avn \)).
    So that a better approximation of the vertex normals can maybe leads to a better order of convergence up to 2.

@@

@@

@@

  \begin{figure}
    \centering\includegraphics[width=.49\textwidth]{bilder/torus/L2K.eps}
    \centering\includegraphics[width=.49\textwidth]{bilder/torus/L2H.eps}
    \caption{Torus \eqref{eqTorus}: Relative \( L_{2} \) error for \( K \) (top) and
                                                     \( H \) (bottom) for different \( h \) in a log-log-plot.}
    \label{figTorusL2Err}
  \end{figure}
